---
title: "RaMP Vignette"
author: "Kyle Spencer and Ewy Mathé"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: yes
    highlight: kate
    theme: yeti
    toc: yes
    fig_width: 9
    fig_height: 7
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Running RaMP locally}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Introduction

This vignette will provide basic steps for interacting with [RaMP](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5876005/) (Relational database of Metabolomic Pathways).  
The codebase for RaMP is available on [our GitHub site](https://github.com/ncats/RaMP-DB).  Details on RaMP installation are also avaialble through GitHub, and questions can be asked through the Issues tab.

This vignette provides the basic steps after RaMP has been installed locally and includes various functionalities:

1) Retrieve Analytes From Input Pathway(s)
2) Retrieve Pathways From Input Analyte(s)
3) Perform Pathway Enrichment
4) Retrieve Analytes Involved in Same Reaction
5) Retrieve Ontologies   

Once installed, first load the package:
```{r}
library(RaMP)
library(formattable) # this is to make some of the outputs look nicer
```

Then set your password for accessing the RaMP DB through MySQL to conpass:
```{r, eval=FALSE}
conpass="password"
```

```{r, echo=FALSE}
conpass=""
```

## Retrieve Analytes From Input Pathway(s)
Analytes (genes, proteins, metabolites) can be retrieve by pathway.  Users have to input the exact pathway name. Here is an example (note that you will need to fill in your password for database access into the parameter "conpass"):
```{r}
myanalytes <- getAnalyteFromPathway(pathway="sphingolipid metabolism",conpass=conpass)
formattable(myanalytes[1:4,])
```

To retrieve information from multiple pathways, input a vector of pathway names:
```{r}
myanalytes <- getAnalyteFromPathway(pathway=c("De Novo Triacylglycerol Biosynthesis", "sphingolipid metabolism"),conpass=conpass)
```

## Retrieve Pathways From Input Analyte(s)
It is oftentimes useful to get a sense of what pathways are represented in a dataset (this is particularly true for metabolomics, where coverage of metabolites varies depending on what platform is used).  In other cases, one may be interested in exploring one or several metabolites to see what pathways they are arepresented in.

Note that it is always preferable to utilize IDs rather then common names.  
When entering IDs, prepend with the database of origin followed by a colon, for example kegg:C02712, hmdb:HMDB04824, etc.. It is possible to input IDs using multiple different sources. RaMP currently supports the following IDs:

```{r}
  con <- RaMP:::connectToRaMP(conpass=conpass)
  query1 <- "select distinct(IDtype) from source where geneOrCompound ='gene';"
  query2 <- "select distinct(IDtype) from source where geneOrCompound ='compound';"
  df1 <- DBI::dbGetQuery(con,query1)
  df2 <- DBI::dbGetQuery(con,query2)
  DBI::dbDisconnect(con)

  print(paste("IDs to be prepended for compounds:",paste(df2[,1],collapse=", ")))
  print(paste("IDs to be prepended for compounds:",paste(df1[,1],collapse=", ")))
```

In this example, we will search for pathways that involve the two genes MDM2 and TP53, and the two metabolites glutamate and creatinine.

```{r} 
pathwaydfids <- getPathwayFromAnalyte(c("ensembl:ENSG00000135679", "hmdb:HMDB0000064",
        "hmdb:HMDB0000148", "ensembl:ENSG00000141510"), conpass=conpass)
formattable(head(pathwaydfids))
```

Note that each row returns a pathway attributed to one of the input analytes.  The last two columns are internal IDs that are not useful to most users and can be ignored.  These last two columns are however needed to run pathway enrichment analysis.
To retrieve the number of unique pathways returned for all analytes or each analyte, try the following:

```{r}
print(paste("Number of Unique Pathways Returned for All Analytes:", length(unique(pathwaydfids$pathwayRampId))))
lapply(unique(pathwaydfids$commonName), function(x) {
        (paste("Number of Unique Pathways Returned for",x,":",
                length(unique(pathwaydfids[which(pathwaydfids$commonName==x),]$pathwayRampId))))})
```

## Perform Pathway Enrichment
Using the pathways that our analytes map to, captured in the pathwaydfids data frame in the previous step, we can now run Fisher's Exact test to identify pathways that are enriched for our analytes of interest:

```{r}
fisher.results <- RaMP::runCombinedFisherTest(pathwaydf = pathwaydfids, conpass=conpass)
```

Once we have our fisher results we can format them into a new dataframe and filter the pathways for significance.
For this example we will be using an FDR p-value cutoff of 0.20.
```{r}
#Returning Fisher Pathways and P-Values
filtered.fisher.results <- FilterFishersResults(fisher.results, p_holmadj_cutoff=0.05)

print("Pathways with Holm-adjusted Pval < 0.2")
print(filtered.fisher.results$fishresults[,c("pathwayName","Pval_combined_Holm")])
```

Because RaMP combines pathways from multiple sources, pathways may be represented more than once.  Further, due to the hierarchical nature of pathways and because Fisher's testing assumes pathways are independent, subpathways and their parent pathways may appear in a list.
To help group together pathways that represent similar biological processes, we have implemented a clustering algorithm that groups pathways together if they share analytes in common.

```{r}
clusters <- RaMP::findCluster(filtered.fisher.results, perc_analyte_overlap = 0.2, 
	min_pathway_tocluster = 3)

# To view the results in a nicer format:
formattable(RaMP:::write_FishersResults(fishResults=clusters))
```

## Retrieve Ontologies from Input Metabolites
RaMP incorporates various types of ontologies as they related to analytes. Here are the ontologies currently found in RaMP: 

```{r echo - FALSE}
con <- RaMP:::connectToRaMP(conpass=conpass)
query1 <- "select * from ontology;"
df1 <- DBI::dbGetQuery(con,query1)
```

```{r echo = FALSE}
#Formatting Output
df1 <- dplyr::select(df1, c("commonName", "biofluidORcellular"))

ontology.table <- matrix(nrow = length(unique(df1$biofluidORcellular)), ncol = ncol(df1))
ontology.table <- as.data.frame(ontology.table)

colnames(ontology.table) <- c("Ontology_Type", "Individual_Ontologies")
ontology.table$Ontology_Type <- unique(df1$biofluidORcellular)

biofluid.vec <- c()
cell.vec <- c()
origins.vec <- c()
tissue.vec <- c()

for(i in 1:nrow(df1)) {
  if(df1$biofluidORcellular[i] == "biofluid") {
    biofluid.vec <- c(biofluid.vec, df1$commonName[i])
  }
  if(df1$biofluidORcellular[i] == "cellular location") {
    cell.vec <- c(cell.vec, df1$commonName[i])
  }
  if(df1$biofluidORcellular[i] == "origins") {
    origins.vec <- c(origins.vec, df1$commonName[i])
  }
  if(df1$biofluidORcellular[i] == "tissue location") {
    tissue.vec <- c(tissue.vec, df1$commonName[i])
  }
}

biofluid.vec <- paste(biofluid.vec, sep = "", collapse = ", ")
cell.vec <- paste(cell.vec, sep = "", collapse = ", ")
origins.vec <- paste(origins.vec, sep = "", collapse = ", ")
tissue.vec <- paste(tissue.vec, sep = "", collapse = ", ")

for(i in 1:nrow(ontology.table)) {
  if(ontology.table$Ontology_Type[i] == "biofluid") {
    ontology.table$Individual_Ontologies[i] <- biofluid.vec
  }
  if(ontology.table$Ontology_Type[i] == "cellular location") {
    ontology.table$Individual_Ontologies[i] <- cell.vec
  }
  if(ontology.table$Ontology_Type[i] == "origins") {
    ontology.table$Individual_Ontologies[i] <- origins.vec
  }
  if(ontology.table$Ontology_Type[i] == "tissue location") {
    ontology.table$Individual_Ontologies[i] <- tissue.vec
  }
}
```

```{r echo=FALSE}
formattable(ontology.table)
```

The function getOntoFromMeta() retrieves ontologies that are associated with a user input list of metabolites. This function takes in a vector of metabolites as an input and the connection information (e.g. password) for access to the RaMP MySQL, and returns a data frame comprised of the ontologies from the user's defined metabolites.

```{r}
metabolites.of.interest <- c("ensembl:ENSG00000135679", "hmdb:HMDB0000064",
        "hmdb:HMDB0000148", "ensembl:ENSG00000141510")

new.ontologies <- RaMP::getOntoFromMeta(analytes = metabolites.of.interest, conpass = conpass)

formattable(head(new.ontologies))
```

## Retrieve Metabolites from Input Ontologies
Conversely, the user can retrieve the metabolites that are associated with a specific ontology or vector of ontologies using the function getMetaFromOnto(). The function input should also include the connection information (e.g. password) for access to the RaMP MySQL. The function will return all metabolites associated with user input ontology(ies).

```{r}
ontologies.of.interest <- c("Colon", "Liver", "Lung")

new.metabolites <- RaMP::getMetaFromOnto(ontology = ontologies.of.interest, conpass = conpass)

formattable(head(new.metabolites))
```

## Retrieve Analytes Involved in the Same Reaction
RaMP tracks metabolites and proteins/genes that are involved in the same reaction. Users can thus retrieve gene transcripts that encode enzymes catalyzing reactions involving metabolites of interest (e.g. altered in a given experiment). 

The function rampFastCata() can be used to this effect. The function input is a vector of metabolites of interest and the connection information (e.g. password) for access to the RaMP MySQL. The user can also input a vector of protein IDs or gene transcripts to return the metabolites involved in chemical reactions with the input proteins or gene transcript encoded proteins.

```{r}
#Input Metabolites
metabolites.of.interest <- c("ensembl:ENSG00000135679", "hmdb:HMDB0000064",
        "hmdb:HMDB0000148", "ensembl:ENSG00000141510")

new.transcripts <- rampFastCata(analytes = metabolites.of.interest, conpass = conpass)

formattable(head(new.transcripts))

#Input Proteins
proteins.of.interest <- c("hmdb:HMDBP00214", "hmdb:HMDBP00703", "kegg:2346")

new.metabolites <- rampFastCata(analytes = proteins.of.interest, conpass = conpass)

formattable(head(new.metabolites))
```

To visualize the output when retrieving analytes involved in the same reaction, RaMP has a built in function, plotCataNetwork(), that generates networks where nodes are analytes and edges denote a "reaction relationship"  This function uses the dataframe created by rampFastCata() as an input. The network plots returned are completely interactive.

```{r}
plotCataNetwork(new.transcripts)
```

## Questions
If you have any questions/suggestions on the functionality of RaMP, please email them at .  For issues/bugs with the code, please create an issue in our GitHub repo (https://github.com/ncats/RaMP-DB). You can also contact us at ncatsramp@mail.nih.gov.



```{r}
sessionInfo()
```